---
title: "Causality"
subtitle: "Econometric Methods"
author: "Zahid Asghar"
date: "06-11-2023"
date-format: "DD-MM-YYYY"
format: 
  revealjs:
    theme: [default, custom.css,styles.scss]
    chalkboard: true
    code-fold: true
    slide-number: c/t
    keep-md: false
execute: 
  freeze: auto
---


```{r, setup, include = F}
 #devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, here, tidyverse,
  emoGG, latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  data.table,
  dplyr,
  lubridate,
  magrittr, knitr, parallel,emoGG
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
```



## Today

Causality

## Upcoming

**Assignments** One assigned; optional assignment coming soon.


# Causality


## Intro

Most tasks in econometrics boil down to one of two goals:

$$
\begin{align}
  y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + u
\end{align}
$$

. . .

1. [Prediction:] Accurately and dependably [predict/forecast]{style=color:purple;} $\color{#6A5ACD}{y}$ using on some set of explanatory variables—doesn't need to be $x_1$ through $x_k$. Focuses on $\color{#6A5ACD}{\hat{y}}$. $\beta_j$ doesn't really matter.

. . .

2. [Causal estimation:]{style=color:purple;}[^1] Estimate the actual data-generating process—learning about the true, population model that explains [how]{style=color:pink;} $\color{#e64173}{y}$ [changes when we change]{style=color:pink;} $\color{#e64173}{x_j}$—focuses on $\color{#e64173}{\beta_j}$. Accuracy of $\hat{y}$ is not important.

[^1]: Often called *causal identification*.


- - -

### For the rest of the term, we will focus on [causally estimating]{style=color:brown;} $\color{#e64173}{\beta_j}$.


## The challenges

### As you saw in the data-analysis exercise, determining and estimating the true model can be pretty difficult—both [practically]{style=color:purple;} and [econometrically]{style=color:purple;}. 


##

::: columns

:::{.column-width=50%}

**Practical challenges**

- Which variables?
- Which functional form(s)?
- Do data exist? How much?
- Is the sample representative?

::: 
:::{.column-width=50%} 

**Econometric challenges**

- Omitted-variable bias
- Reverse causality
- Measurement error
- How precise can/must we be?
:::
:::

## 

Many of these challenges relate to [exogeneity]{style=color:purple;}, _i.e._, $\color{#314f4f}{\mathop{\boldsymbol{E}}\left[ u_i | X \right] = 0}$.

<br>Causality requires us to [hold all else constant]{style=color:purple;} (*ceterus paribus*).


### It's complicated

Occasionally, [*causal*]{style=color:purple;} relationships are simply/easily understood, _e.g._,


- What [caused]{style=color:purple;} the forest fire?
- [How]{style=color:purple;} did this baby get here?



Generally, [*causal*]{style=color:purple;} relationships are complex and challenging to answer, _e.g._,

- - -

- What [causes]{style=color:purple;} some countries to grow and others to decline?
- What [caused]{style=color:purple;} the capital riot?
- Did lax regulation [cause]{style=color:purple;} Texas's recent energy problems?
- [How]{style=color:purple;} does the number of police officers affect crime?
- What is the [effect]{style=color:purple;} of better air quality on test scores?
- Do longer prison sentences [decrease]{style=color:purple;} crime?
- How did cannabis legalization [affect]{style=color:purple;} mental health/opioid addiction?


## Correlation ≠ Causation

You've likely heard the saying

> Correlation is not causation.

The saying is just pointing out that there are violations of exogeneity.

. . .

Although correlation is not causation, [causation *requires* correlation]{style=color:purple;}.

. . .

[New saying:]{style=color:purple;}

> Correlation plus exogeneity is causation.


Let's work through a few examples.

# Causation



## Example: The causal effect of fertilizer[^2]


[^2] Many of the early statistical and econometric studies involved agricultural field trials.


Suppose we want to know the causal effect of fertilizer on corn yield.



**Q:** Could we simply regress yield on fertilizer? 

. . .

<br>**A:** Probably not (if we want the causal effect).

. . . 

<br><br>**Q:** Why not? 

- - - 

<br>**A:** Omitted-variable bias: Farmers may apply less fertilizer in areas that are already worse on other dimensions that affect yield (soil, slope, water).<br>[Violates *all else equal* (exogeneity). Biased and/or spurious results.]{style=color:purple;}

. . . 

<br><br>**Q:** So what *should* we do?
. . . 

<br>**A:** [Run an experiment!]{style=color:red;}  💩


## Example: The causal effect of fertilizer

Randomized experiments help us maintain *all else equal* (exogeneity).



We often call these experiments [*randomized control trials*]{style=color:purple;} (RCTs)..super[^3]

.footnote[
[^3] Econometrics (and statistics) borrows this language from biostatistics and pharmaceutical trials.
]

. . .

Imagine an RCT where we have two groups:

- [Treatment:]{style=color:purple;} We apply fertilizer.
- [Control:]{style=color:pink;} We do not apply fertilizer.

By randomizing plots of land into [treatment]{style=color:purple;} or [control]{style=color:purple;}, we will, on average, include all kinds of land (soild, slope, water, *etc.*) in both groups.

##

*All else equal*!

[54 equal-sized plots]{style=color:purple;}
```{r, fertilizer_plot1, echo = F}
xd <- 9
yd <- 6
set.seed(123)
fert_df <- expand.grid(x = 1:xd, y = 1:yd) %>%
  mutate(
    q = x + y + 1.25 * rnorm(xd*yd),
    trt = sample(x = c(T, F), size = xd*yd, replace = T, prob = c(0.51, 0.49)),
  ) %>%
  arrange(-y, x) %>%
  mutate(id = 1:(xd*yd) %>% str_pad(2, "left", "0")) %>%
  arrange(y, x)
ggplot(data = fert_df, aes(x, y)) +
geom_tile(color = "grey40", fill = "grey93", size = 0.2) +
geom_text(aes(label = id), color = "grey45", size = 5, family = "Roboto Mono") +
theme_void() +
coord_equal()
```


##

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;}
```{r, fertilizer_plot2, echo = F}
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_1, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 1), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```


##

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_2, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 2), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_3, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 3), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

##

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_4, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 4), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_5, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 5), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_6, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 6), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_7, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 7), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

##

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_8, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 8), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_9, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 9), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_10, echo = F}
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## 

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_11, echo = F}
set.seed(248)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

##

[54 equal-sized plots]{style=color:purple;} [of varying quality]{style=color:orange;} [plus randomly assigned treatment]{style=color:purple;}
```{r, fertilizer_plot3_12, echo = F}
set.seed(250)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## Example: The causal effect of fertilizer

We can estimate the [causal effect]{style=color:purple;} of fertilizer on crop yield by comparing the average yield in the treatment group (💩) with the control group (no 💩).

$$
\begin{align}
  \overline{\text{Yield}}_\text{Treatment} - \overline{\text{Yield}}_\text{Control}
\end{align}
$$

. . .

Alternatively, we can use the regression

. . .

$$
\begin{align}
  \text{Yield}_i = \beta_0 + \beta_1 \text{Trt}_i + u_i \tag{1}
\end{align}
$$

where $\text{Trt}_i$ is a binary variable (=1 if plot $i$ received the fertilizer treatment).

. . .

**Q:** Should we expect $(1)$ to satisfy exogeneity? Why? 

. . . 

**A:** On average, [randomly assigning treatment should balance]{style=color:purple;} trt. and control across the other dimensions that affect yield (soil, slope, water).



## Example: Returns to education

Labor economists, policy makers, parents, and students are all interested in the (monetary) *return to education.*

. . .

[Thought experiment:]{style=color:purple;}
- Randomly select an individual.
- Give her an additional year of education.
- How much do her earnings increase?

This change in earnings gives the [causal effect]{style=color:purple;} of education on earnings.

## 


**Q:** Could we simply regress earnings on education?

<br>**A:** Again, probably not if we want the true, causal effect.


1. People *choose* education based upon many factors, *e.g.*, ability.
1. Education likely reduces experience (time out of the workforce).
1. Education is [*endogenous*]{style=color:purple;} (violates *exogeneity*).


The point (2) above also illustrates the difficulty in learning about educations while *holding all else constant*.

Many important variables have the same challenge—gender, race, income.

## 
**Q:** So how can we estimate the returns to education?

::: {.columns}
::: {.column width='50%'}
[Option 1:]{style=color:purple;} Run an [experiment]{style=color:orange;}.

- Randomly [assign education]{style=color:orange;} (might be difficult).
- Randomly [encourage education]{style=color:pink;} (might work).
- Randomly [assign programs]{style=color:orange;} that affect education (*e.g.*, mentoring).
:::
::: {.column width='50%'}
[Option 2:]{style=color:pink;} Look for a [*natural experiment*]{style=color:purple;} 

— a policy or accident in society that arbitrarily increased education for one subset of people.

- Admissions [cutoffs]{style=color:pink;}
- [Lottery]{style=color:purple;} enrollment and/or capacity [constraints]{style=color:pink;}
:::
:::

## Real-world experiments

Both examples consider [real experiments]{style=color:pink;} that isolate causal effects.

[Characteristics]{style=color:pink;}

- [Feasible]{style=color:purple;}—we can actually (potentially) run the experiment.
- [Compare individuals]{style=color:purple;} randomized into treatment against individuals randomized into control.
- [Require "good" randomization]{style=color:purple;} to get *all else equal* (exogeneity).

. . .

*Note:* Your experiment's results are only as good as your randomization.

## 

[Unfortunate randomization]{style=color:purple;}
```{r, fertilizer_plot3_bad, echo = F}
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = filter(fert_df, x < xd/3 | y <= yd/3 | (x == xd & y == yd)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

## The ideal experiment

The [ideal experiment]{style=color:purple;} would be subtly different.

Rather than comparing units randomized as [treatment]{style=color:purple;} vs. [control]{style=color:purple;}, the ideal experiment would compare treatment and control [for the same, exact unit]{style=color:purple;}.

. . .

$$
\begin{align}
  y_{\text{Treatment},i} - y_{\text{Control},i}
\end{align}
$$

. . .

which we will write (for simplicity) as

$$
\begin{align}
  y_{1,i} - y_{0,i}
\end{align}
$$

. . .

This [[*ideal experiment*]{style=color:purple;} is clearly infeasible.][^4], but it creates nice notation for causality (the Rubin causal model/Neyman potential outcomes framework).

[^4]: Without (1) God-like abilities and multiple universes or (2) a time machine.


##

::: {.columns}

:::{.column width='50%'}


The *ideal* data for 10 people
```{r, ideal_data, echo = F}
set.seed(3)
ideal_df <- data.frame(
  i = 1:10,
  trt = rep(c(1, 0), each = 5),
  y1i = c(runif(10, 4, 10) %>% round(2)),
  y0i = c(runif(10, 0, 5) %>% round(2))
)
ideal_df
```

:::

::: {.column width='50%'}


Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.
:::
:::: 

- - - 

::: {.columns}

:::{.column width='50%'}
The *ideal* data for 10 people
```{r, ideal_data_trt, echo = F}
ideal_df %>% mutate(effect_i = y1i - y0i)
```
:::



:::{.column width='50%'}

Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.
:::
::::

- - - 

::: {.columns}

:::{.column width='50%'}
The *ideal* data for 10 people
```{r, ideal_data_trt2, echo = F}
ideal_df %>% mutate(effect_i = y1i - y0i)
```
:::



:::{.column width='50%'}


Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.

The mean of $\tau_i$ is the<br>[average treatment effect]{style=color:purple;} ([ATE]{style=color:purple;}).

Thus, $\color{#e64173}{\overline{\tau} = `r transmute(ideal_df, effect_i = y1i - y0i) %>% unlist %>% mean %>% round(2)`}$

:::
::::

- - -

This model highlights the fundamental problem of causal inference.
$$
\begin{align}
  \tau_i = \color{#e64173}{y_{1,i}} &- \color{#6A5ACD}{y_{0,i}}
\end{align}
$$

. . .

[The challenge:]{style=color:purple;}

If we observe $\color{#e64173}{y_{1,i}}$, then we cannot observe $\color{#6A5ACD}{y_{0,i}}$.
<br>If we observe $\color{#6A5ACD}{y_{0,i}}$, then we cannot observe $\color{#e64173}{y_{1,i}}$.

##


So a dataset that we actually observe for 6 people will look something like 

::: {.columns}

:::{.column width='50%'} 

```{r, ideal_data_obs, echo = F}
obs_df <- ideal_df
obs_df$y0i[1:5] <- NA
obs_df$y1i[6:10] <- NA
obs_df
```

:::
:::{.column width='50%'}

We can't observe $\color{#e64173}{y_{1,i}}$ and $\color{#6A5ACD}{y_{0,i}}$.

But, we do observe 

- $\color{#e64173}{y_{1,i}}$ for $i$ in 1, 2, 3, 4, 5 

- $\color{#6A5ACD}{y_{0,j}}$ for $j$ in 6, 7, 8, 9, 10

:::
::::


**Q:** How do we "fill in" the `NA`s and estimate $\overline{\tau}$?

## Causally estimating the treatment effect


[Notation:]{style=color:purple;} Let $D_i$ be a binary indicator variable such that

- $\color{#e64173}{D_i=1}$ [if individual]{style=color:purple;} $\color{#e64173}{i}$ [is treated]{style=color:purple;}.
- $\color{#6A5ACD}{D_i=0}$ [if individual]{style=color:purple;} $\color{#6A5ACD}{i}$ [is not treated (*control* group).]{style=color:purple;}

. . .

Then, rephrasing the previous slide,

- We only observe $\color{#e64173}{y_{1,i}}$ when $\color{#e64173}{D_{i}=1}$.
- We only observe $\color{#6A5ACD}{y_{0,i}}$ when $\color{#6A5ACD}{D_{i}=0}$.

. . .

**Q:** How can we estimate $\overline{\tau}$ using only $\left(\color{#e64173}{y_{1,i}|D_i=1}\right)$ and $\left(\color{#6A5ACD}{y_{0,i}|D_i=0}\right)$?

- - -

**Q:** How can we estimate $\overline{\tau}$ using only $\left(\color{#e64173}{y_{1,i}|D_i=1}\right)$ and $\left(\color{#6A5ACD}{y_{0,i}|D_i=0}\right)$?

. . .

**Idea:** What if we compare the groups' means? _I.e._,
$$
\begin{align}
  \color{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}
\end{align}
$$

. . .

**Q:** When does this simple difference in groups' means provide information on the [causal effect]{style=color:purple;} of the treatment?

. . .

**Q.sub[2.0]:** Is $\color{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\overline{\tau}$?

. . . 

Time for math! [🎉]{width=200}

- - -

[Assumption:]{style=color:purple;} Let $\tau_i = \tau$ for all $i$.

This assumption says that the treatment effect is equal (constant) across all individuals $i$.

. . .

[Note:]{style=color:purple;} We defined

$$
\begin{align}
  \tau_i = \tau = \color{#e64173}{y_{1,i}} - \color{#6A5ACD}{y_{0,i}}
\end{align}
$$

which implies

$$
\begin{align}
   \color{#e64173}{y_{1,i}} = \color{#6A5ACD}{y_{0,i}} + \tau
\end{align}
$$

## 

**Q.sub[3.0]:** Is $\color{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\tau$?

. . .

Difference in groups' means

<br> $\quad \color{#ffffff}{\Bigg|}=\color{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$
. . .
<br> $\quad \color{#ffffff}{\Bigg|}=\color{#e64173}{\mathop{Avg}\left( y_{1,i}\mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$
. . . 
<br> $\quad \color{#ffffff}{\Bigg|}=\color{#e64173}{\mathop{Avg}\left( \color{#000000}{\tau \: +} \: \color{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$
. . .
<br> $\quad \color{#ffffff}{\Bigg|}=\tau + \color{#e64173}{\mathop{Avg}\left(\color{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \color{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$
. . .
<br> $\quad \color{#ffffff}{\Bigg|}= \text{Average causal effect} + \color{#FFA500}{\text{Selection bias}}$

##

So our proposed group-difference estimator give us the sum of

1. $\tau$, the [causal, average treatment effect]{style=color:purple;} that we want
2. [Selection bias:]{style=color:orange;} How much trt. and control groups differ (on average).
